### 2022-10-18 manuscript/manuscript.tex
more theory

After meiosis, a fraction $1-r$ of these associations between the tracked
neutral neutral allele $x_{0,1}'$ and fitness background $f_{0,i}'$ persist,
and a fraction $r$ disassociate its currently coupled background and create an
association with the homologous fitness background, $f_{0,i}''$. When linkage
is tight, this latter term can be ignored as we will do here, though it does
impact background levels of neutral diversity \parencite{Santiago1995-hx}.
Additionally, the fitness effects of the selected background may change in
later generations, in complex ways \parencite{Barton1986-yh,Turelli1990-kd}.
\textcite{Santiago1995-hx} assume an equilibrium level of fitness variation
$V_h$, with the fitness variance associated with a particular background
decaying at a simple geometric decay at rate $Z$ each generation, which is
sufficient for background selection. An approximate rate of the decay can be
worked out for the particular selective system. In general, this can be a much
more complicated function; \textcite{Buffalo2019-qs} show that even fluctuating
selection can be accommodated.

@@

\begin{align}
  S_{2}' &= S_1 Z(1-r) \\
  S_{3}' &= S_1 Z^2(1-r)^2 \\
  S_{4}' &= S_1 Z^3(1-r)^3.
\end{align}

@@

\begin{align}
  S_1 Q_t &= S_1 + S_2' + S_3' + \ldots + S_t' \\
  Q_t &= 1 + \sum_{i=1}^t r^i (1-Z)^i \\
  Q_\infty &= \frac{1}{1-r (1-Z)}
\end{align}

@@

\begin{align}
  \E(S_1 S_2) &= \E(S_1^2) Z(1-r) = \E(S_1^2) Q_2^2 \\
  \E(S_1 S_3) &= \E(S_1^2) Z^2(1-r)^2 = \E(S_1^2) Q_3^2.
\end{align}

@@

The results of the previous section assume only a single locus a recombination
fraction $r$ apart from the neutral locus is determining the coalesce rate. The
model of \textcite{Santiago1998-bs} extends this to the case where fitness is
polygenic. As they show in their original paper, this system can accommodate a
variety of different equilibrium selection systems as long as expressions for
$V_h$ and $Z$ can be worked out. Other authors have extended similar models to
more elaborate bbreeding structures \parencite{Wray1990-zf,Woolliams1993-qo}.

@@

(c.f. \cite{Santiago1998-bs} equation 4b). This summation can then in turn be
approximated by a integral; see \textcite{Santiago1998-bs} for more. Note that
another way to arrive at this equation is to take the product of reduction
factors $\nicefrac{N_e}{N}$ across selected sites, as in equation 4 of
\textcite{Hudson1995-xc}.

The heritable variance in fitness $V_h$ associated with the tracked neutral
allele does not remain constant through the generations, but decays due to
selection and drift, and is maintained by the input of new mutations. For an
arbitrary polygenic selective system, the dynamics are extraordinarily
complicated due to the complexity of multilocus selection. However, several
quantitative genetic models simplify these dynamics by assuming an
infinitesimal model, where the trait is sufficiently polygenic that the allele
frequency changes at the underlying loci due to selection are negligible
compared to drift. This allows for a critical assumption of the quantitative
genetic models of linked selection, that the decay rate in variation is
constant (and thus independent of the underlying allele frequency changes). In
the Santiago and Caballero model, this is the decay factor $Z$.

The theory of \textcite{Santiago1998-bs} is based on an equilibrium model,
where $\Delta V_h = 0$ of the long run, even if the fitness variation
\emph{associated} with the allele's haplotype is changing. This macro-scale

\begin{align}
  \Delta V_h &= V_m - V_h\left(1 - \frac{1}{2 N_e}\right) \\
\end{align}

@@

Classic background selection theory assumes an infinite population size, such
that mutation frequencies are at their mutation-selection equilibrium and the
expected number of mutations is $n_i = \nicefrac{\mu}{s}L$ where $L$ is the
total number of basepairs that can be mutated (the mutational target size).
Under a multiplicative fitness function in an infinite population, selection
cannot generate linkage disequilibria \parencite{Turelli1990-kd}. Thus the
classic strong background selection model assumes additive genetic fitness is
equal to the additive genic fitness (which excludes the contribution of
covariance between selected sites),

\begin{align}
  \V_i(w_i) &= 2 \sum_{l=1}^L p(1-p) s^2 \\
            &= 2 L \frac{\mu}{s} \left( 1 - \frac{\mu}{s}\right) s^2 = U s + \mathcal{O}(\mu^2) \approx Us.
\end{align}

where we define $U := 2 L \mu$ to be the deleterious mutation rate per diploid
genome, per generation.

Next, \textcite{Santiago1998-bs} find a general
equation for the factor $Z$ by which fitness variation associated with the
neutral allele is reduced under mutation-selection balance. This is a
macro-scale variable reflecting the average balance between the average loss of
fitness variance due to selection under an infinitesimal model and drift. Each
generation, the gain of new fitness variance per generation is $V_m$.

\begin{align}
  Z = 
\end{align}

Then, with under the strong background selection model,
a site under purifying selection reduces diversity at a focal neutral site $r$
by a factor of

\begin{align}
  \frac{N_{e,t}}{N} = \exp \left(-\frac{V_{h,1} Q_{t,1}^2}{2}\right)
\end{align}

% \begin{align}
%   \frac{\var(x_t - p_0)}{p_0(1-p_0)} &= \sum_{i=1}^\infty \left(\frac{V_h Q_i^2}{2 N_{e,i}} + \frac{V_n}{8N_{e,i}} + \frac{1}{4N_{e,i}} \right)\prod_{t=1}^{i-1} \left(1-\frac{1}{2 N_{e,t}}\right) \\
% \end{align}

>>>>

After meiosis, a fraction $1-r$ of these associations between the tracked
neutral neutral allele $x_{0,1}'$ and fitness background $f_{0,i}'$ persist,
and a fraction $r$ disassociate its currently coupled background and create an
association with the homologous fitness background, $f_{0,i}''$. When linkage
is tight, this latter term can be ignored as we will do here, though it does
impact background levels of neutral diversity \parencite{Santiago1995-hx}.
Additionally, the fitness effects of the selected background may change in
later generations, in complex ways \parencite{Barton1986-yh,Turelli1990-kd}.
\textcite{Santiago1995-hx} assume an equilibrium level of fitness variation
$V_h$, with the fitness variance associated with a particular background
decaying at a simple geometric decay at rate $1-\kappa$ each generation, which
is sufficient for background selection. An approximate rate of the decay can be
worked out for the particular selective system. In general, this can be a much
more complicated function; \textcite{Buffalo2019-qs} show that even fluctuating
selection can be accommodated.

@@

\begin{align}
  S_{2}' &= S_1 (1-\kappa)(1-r) \\
  S_{3}' &= S_1 (1-\kappa)^2(1-r)^2 \\
  S_{4}' &= S_1 (1-\kappa)^3(1-r)^3.
\end{align}

@@

\begin{align}
  \label{eq:qterm}
  S_1 Q_t &= S_1 + S_2' + S_3' + \ldots + S_t' \\
  Q_t &= 1 + \sum_{i=1}^t \left((1-r)\kappa\right)^i \\
  Q_\infty &= \frac{1}{1-(1-r) \kappa}
\end{align}

@@

\begin{align}
  \E(S_1 S_2) &= \E(S_1^2) (1-\kappa)(1-r) = \E(S_1^2) Q_2^2 \\
  \E(S_1 S_3) &= \E(S_1^2) (1-\kappa)^2(1-r)^2 = \E(S_1^2) Q_3^2.
\end{align}

@@

The results of the previous section assume only a single locus a recombination
fraction $r$ apart from the neutral locus is determining the coalesce rate. The
model of \textcite{Santiago1998-bs} extends this to the case where fitness is
polygenic. As they show in their original paper, this system can accommodate a
variety of different equilibrium selection systems as long as expressions for
$V_h$ and $1-\kappa$ can be worked out. Other authors have extended similar models to
more elaborate bbreeding structures \parencite{Wray1990-zf,Woolliams1993-qo}.

@@

(c.f. \cite{Santiago1998-bs} equation 4b). Santiago and Caballero ignore the
covariances between the locus-specific variation decay rates $1-\kappa_l$ terms
and the locus-specific variances, $V_{h,l}$. This allows for the terms to be
separated,

\begin{align}
  N_{e,t} &\approx N \exp\left(- \frac{V_h}{2} \sum_{l=1}^L Q_{t,l}^2\right).
\end{align}

\subsection*{Continuous Approximation to the Segment Under Selection}

Our model differs from that of \textcite{Santiago1998-bs} in that we do not
integrate over the entire genome. In their 1998 model, Santiago and Caballero
consider the impact of both sites under selection on the same chromosome, as
well as sites on independently assorting chromosomes. We only consider the
contribution of a single segment under purifying selection, and the reduction
from additional segments accumulate multiplicatively (similar to our
implementation of the classic BGS model, XXX).

If we assume that the $L$ selected sites are evenly distributed in a region of
length $G$ basepairs, we can rewrite the sum as $\nicefrac{L}{G} \sum_{l=1}^G
Q_{t,l}^2$

\begin{align}
  N_{e,t} &\approx N \exp\left(- \frac{V_h}{2} \sum_{l=1}^L Q_{t,l}^2\right).
\end{align}

For a continuous recombination map, the recombination fraction between the
neutral allele and a selected site $d$ map distance away is given by Haldane's
map function, $r = (1-e^{-2d})/2$. Santiago and Caballero assume the focal
neutral site is in the middle of the chromosome, and integrate over the map
distance. Assuming that linkage is not too tight,

\begin{align}
  Q_\infty &= \frac{1}{1-r \kappa}
\end{align}

For our model, we consider the contribution of individual contiguous segments
of basepairs rather than the whole genome.

This summation can then in turn be
approximated by a integral; see \textcite{Santiago1998-bs} for more. Note that
another way to arrive at this equation is to take the product of reduction
factors $\nicefrac{N_e}{N}$ across selected sites, as in equation 4 of
\textcite{Hudson1995-xc}.

\subsection*{The Variance Dynamics}

The heritable variance in fitness $V_h$ associated with the tracked neutral
allele does not remain constant through the generations, but decays due to
selection and drift, and is maintained by the input of new mutations. For an
arbitrary polygenic selective system, the dynamics are extraordinarily
complicated due to the complexity of multilocus selection, and the dynamics of
the trait response to selection depend on the variance, which in turn depends
on higher moments of the fitness distribution. We discuss the simplification of
Santiago and Caballero's model here; hereafter, we assume that $V_h = V_A$, the
genetic variance in additive fitness put on a relative scale such that
$\E_i(f_i) = 1$.

Each generation, the genetic variance changes due to changes in the allele
frequencies at selected sites and linkage disequilibria between these sites.
These are typically expressed as recursions, but certain models permit these
changes to be expressed as a proportional reduction (e.g. truncation selection
and stabilizing selection models; \cite{Keightley1988-eq} p. 36, and
\cite{Walsh2018-bt} p. 557). Santiago and Caballero's model follows this
approach, assuming the variance decays due to selection and drift at a constant
rate $Z := 1-\kappa$ through the generations, such that the variance in the next
generation excluding mutation is $V_A' = (1-\kappa) V_A$ and $\Delta V_A =
-\kappa V_A$.

With mutation, $\Delta V_A = V_m - \kappa V_A$. The model of Santiago and
Caballero assumes that under the long-run equilibrium, $\Delta V_A = 0$, such
that $\kappa V_A = V_m$, and thus the rate of loss is $\kappa =
\nicefrac{V_m}{V_A}$. Then the variance for the dynamics simplify to,

\begin{align}
  \label{eq:Z}
  V_A' &= \left(1 - \frac{V_m}{V_A}\right) V_A \\
  Z &:= \frac{V_A'}{V_A} = 1 - \kappa = 1 - \frac{V_m}{V_A}
\end{align}

(c.f. \cite{Santiago1998-bs} equation 9).

% TODO this is a good description that could make its way to maintext

@@

Classic background selection theory assumes an infinite population size, such
that mutation frequencies are at their mutation-selection equilibrium and the
expected number of mutations across individuals is $\E_i(n_i) =
\nicefrac{\mu}{s}L$ where $L$ is the total number of basepairs that can be
mutated (the mutational target size).  Under a multiplicative fitness function
in an infinite population, selection cannot generate linkage disequilibria
\parencite{Turelli1990-kd}. Thus the classic strong background selection model
assumes additive genetic fitness is equal to the additive genic fitness (which
excludes the contribution of covariance between selected sites),

\begin{align}
  V_\text{BGS} &:= \V_i(w_i) = 2 \sum_{l=1}^L p(1-p) s^2 \\
            &= 2 L \frac{\mu}{s} \left( 1 - \frac{\mu}{s}\right) s^2 = U s + \mathcal{O}(\mu^2) \approx Us.
\end{align}

where $U$ is defined $U := 2 L \mu$, the deleterious mutation rate per diploid
genome, per generation. They then set $V_A = V_\text{BGS}$, and note that $V_m
\approx Us^2$, since mutations have frequency of $\nicefrac{1}{2N}$ and each
increase the variance by a factor of $s^2$. Note that because the classic BGS
model assumes deterministic selection dynamics (i.e. an infinite population),
the equilibrium $Z$ given by equation \eqref{eq:Z} excludes the reduction due
to drift, which reduces the variance by an additional factor $1 -
\nicefrac{1}{2N_e}$. Thus, the reduction factor under the strong BGS model is,

\begin{align}
  Z \left(1 - \frac{1}{2N_e}\right) = (1 - s)\left(1 - \frac{1}{2N_e}\right) \approx 1 - s -\frac{1}{2N_e}
\end{align}

(c.f. \cite{Santiago1998-bs} equation 10). An alternate derivation can be found
by working out the deterministic reduction in variance from the single-locus
dynamics. Thus, $Q' = $

the reduction under the classic BGS model is,

\begin{align}
  N_e = N \exp()
\end{align}

\subsection*{Weak Selection Dynamics}

With the possibility of weak selection, deleterious alleles can drift up in
frequency and fix. Consequently, the fitness distribution is no longer
stationary, as the set of haplotypes without any deleterious mutations can be
lost due to drift. The approach of \textcite{Santiago2016-mu} that we implement
in our methods is to derive an equation for
